A dynamic bioenergetic model for coral-\emph{Symbiodinium} symbioses and coral bleaching as an alternate stable state

### 2017-02-07 JTB_ms/JTB_ms.tex
remove kahako (causing latex error)

\usepackage[T1]{fontenc}
\usepackage{lmodern}
\usepackage{amssymb,amsmath}
\usepackage{ifxetex,ifluatex}
\usepackage{fixltx2e} % provides \textsubscript
% use upquote if available, for straight quotes in verbatim environments
\IfFileExists{upquote.sty}{\usepackage{upquote}}{}
\ifnum 0\ifxetex 1\fi\ifluatex 1\fi=0 % if pdftex
  \usepackage[utf8]{inputenc}
\else % if luatex or xelatex
  \usepackage{fontspec}
  \ifxetex
    \usepackage{xltxtra,xunicode}
  \fi
  \defaultfontfeatures{Mapping=tex-text,Scale=MatchLowercase}
  \newcommand{\euro}{€}
\fi
% use microtype if available
\IfFileExists{microtype.sty}{\usepackage{microtype}}{}
\usepackage{longtable}
\usepackage{graphicx}
% We will generate all images so they have a width \maxwidth. This means
% that they will get their normal width if they fit onto the page, but
% are scaled down if they would overflow the margins.
\makeatletter
\def\maxwidth{\ifdim\Gin@nat@width>\linewidth\linewidth
\else\Gin@nat@width\fi}
\makeatother
\let\Oldincludegraphics\includegraphics
\renewcommand{\includegraphics}[1]{\Oldincludegraphics[width=\maxwidth]{#1}}
\ifxetex
  \usepackage[setpagesize=false, % page size defined by xetex
              unicode=false, % unicode breaks when used with xetex
              xetex]{hyperref}
\else
  \usepackage[unicode=true]{hyperref}
\fi
\hypersetup{breaklinks=true,
            bookmarks=true,
            pdfauthor={},
            pdftitle={A simplified dynamic bioenergetic model for coral-Symbiodinium symbioses and coral bleaching as an alternate stable state},
            colorlinks=true,
            urlcolor=blue,
            linkcolor=magenta,
            pdfborder={0 0 0}}
\urlstyle{same}  % don't use monospace font for urls
\setlength{\parindent}{0pt}
\setlength{\parskip}{6pt plus 2pt minus 1pt}
\setlength{\emergencystretch}{3em}  % prevent overfull lines
\setcounter{secnumdepth}{0}
% Pandoc toggle for numbering sections (defaults to be off)
\setcounter{secnumdepth}{0}
% Pandoc header

@@

  \title{A simplified dynamic bioenergetic model for coral-\emph{Symbiodinium}
symbioses and coral bleaching as an alternate stable state}
    \author[University of Hawaii]{Ross Cunning\corref{c1}}
   \ead{ross.cunning@gmail.com} 
   \cortext[c1]{Corresponding Author}
    \author[University of California]{Erik B. Muller}

    \author[University of Hawaii]{Ruth D. Gates}

    \author[University of California]{Roger M. Nisbet}

      \address[University of Hawaii]{Hawaii Institute of Marine Biology, Kaneohe, HI 96744, USA}
    \address[University of California]{Department of Ecology, Evolution, and Marine Biology, Santa Barbara, CA
93106, USA}

  \begin{abstract}
  Coral reef ecosystems owe their ecological success--and vulnerability to
  climate change--to the symbiotic metabolism of corals and
  \emph{Symbiodinium} spp. The urgency to understand and predict the
  stability and breakdown of these symbioses (i.e., coral `bleaching')
  demands the development and application of theoretical tools; however,
  the complexity of some approaches may limit their potential application
  and utility. Here, we develop a simplified dynamic bioenergetic model of
  coral-\emph{Symbiodinium} symbioses that demonstrates realistic
  steady-state patterns in coral growth and symbiont abundance across
  gradients of light, nutrients, and feeding. Furthermore, through a
  mechanistic treatment of photo-oxidative stress, the model displays
  dynamic behaviors of bleaching and recovery, and suggests that these
  phenomena represent transitions between alternate stable states. A suite
  of complex responses to multiple, interacting environmental factors
  reproduced by the model suggests it unifyingly captures many important
  attributes of the system; meanwhile, its modular framework and open
  source R code are designed to facilitate expansion and problem-specific
  applications. We see significant potential applications for this
  modeling framework in generating testable hypotheses and predicting
  integrated, mechanistic responses of corals to environmental change.
  \end{abstract}

@@

In reef coral symbioses, intracellular \emph{Symbiodinium} translocate
photosynthetically fixed carbon to support coral metabolism, while the
animal host provides access to inorganic nutrients and carbon dioxide
(Muscatine and Porter 1977). Previous application of DEB theory to this
syntrophic system (Muller et al. 2009) demonstrated a stable symbiotic
relationship and qualitatively realistic growth and biomass ratios
across gradients of ambient irradiance, nutrients, and food. This model
assumed that 1) \emph{Symbiodinium} has priority access to fixed carbon
through photosynthesis, 2) the coral animal has priority access to
dissolved nitrogen through contact with seawater, and 3) each partner
shares with the other only what it cannot use for its own growth. In its
simplest form, this principle of sharing the surplus is sufficient to
describe the dynamics of diverse syntrophic organs and organisms (e.g.,
trees, duckweeds, corals), suggesting the mechanism is mathematically
and evolutionarily robust (Nisbet et al., in prep.).

@@

\begin{enumerate}
\def\labelenumi{\arabic{enumi}.}
\item
  \emph{Develop a module of photooxidative stress.} Of primary interest
  to coral biologists and ecologists is symbiosis dysfunction under
  environmental stress, resulting in coral ``bleaching''--the loss of
  algal symbionts from the association (Jokiel and Coles 1977).
  Photooxidative stress in \emph{Symbiodinium} is considered a primary
  trigger of bleaching in response to high temperature and/or light
  (Weis 2008), and prolonged or severe bleaching can result in
  mortality, though corals sometimes recover their symbionts. Bleaching
  susceptibility, severity, and recovery may be influenced by
  interacting factors such as heterotrophy and nutrient availability
  (Wooldridge 2014b). To simulate these events and interactions, we
  develop a photooxidative stress module linking overreduction of the
  photosynthetic light reactions to downstream impacts of
  photoinhibition and photodamage.
\item
  \emph{Reduce theoretical and mathematical complexity.} Following the
  logic of Jager, Martin, and Zimmer (2013), we exclude certain features
  of formal DEB theory in order to capture behaviors of interest with
  the simplest possible formulation. Here, we present a model without
  reserves, maturity, or reproduction (see Kooijman 2010). This
  formulation restricts the model's scope to the bioenergetics of adult
  corals (i.e., reproduction, larval stages, and metamorphosis are not
  considered), but greatly reduces theoretical complexity and parameter
  numbers, which is advantageous given the relative paucity of data for
  corals. However, our primary motivation for reducing complexity was to
  increase accessibility and applicability for biologists and ecologists
  without requiring significant expertise in DEB theory.
\item
  \emph{Provide well-documented, open-access code.} In order to
  facilitate the continued development and application of theoretical
  modeling tools for coral symbioses, we provide open access to the
  model in the form of detailed, commented code written in the R
  language (R Core Team 2014). With an accessible and modular framework,
  we envision this model as a resource for futher development by the
  scientific community to include additional complexity and
  problem-specific components. We chose the R language, which is freely
  available and in common use by biologists and ecologists, to reach the
  widest possible audience with this work.
\end{enumerate}

@@

In this dynamical system, both partners acquire and use carbon and
nitrogen to construct biomass: the symbiont fixes carbon through
photosynthesis, and receives nitrogen shared by the host, while the host
acquires nitrogen from the environment and receives carbon shared by the
symbiont. A graphical representation of the model is presented in Fig.
1, and each model flux and parameter is defined in Tables 1 and 2,
respectively. We use C-moles as the unit of biomass for consistency with
the rigorous mass balance of DEB theory: 1 C-mole is equivalent to the
amount of biomass containing 1 mole of carbon atoms. Host biomass
(\(H\)), symbiont biomass (\(S\)), and prey biomass (\(X\)) have fixed,
but different, molar N:C ratios (Table 2). Biomass is produced from
carbon and nitrogen by synthesizing units (SU), which are mathematical
specifications of the formation of a product from two substrates; we use
the parallel complementary formulation of Kooijman (2010) to specify
these fluxes. The two state variables of this system are symbiont
biomass and coral biomass; because resources are acquired proportionally
to surface area, and surface area is proportional to volume (i.e.,
corals are ``V1-morphs'' in DEB terminology (Kooijman 2010)), biomass
increases exponentially during growth (indeed, corals grow exponentially
(Bak 1976)).

Environmental stress is implemented in the form of photooxidative
stress, which is thought to be the primary trigger of coral bleaching
(Lesser 1997; Weis 2008; Wooldridge 2009). To simulate these events, we
model the absorption and quenching of light energy by photochemistry and
non-photochemical quenching, and the phenomenological responses that
occur when these capacities are overwhelmed (i.e., photoinhibition,
photodamage, and symbiont loss). While bleaching in response to high
light alone has been observed experimentally (Schutter et al. 2011;
Downs et al. 2013), mass coral bleaching events occur in response to
high temperature (Hoegh-Guldberg 1999); thus, it is important to justify
our consideration of light as the primary stressor. In reality, light
and temperature interact synergistically (Coles and Jokiel 1978; Jones
et al. 1998), and in fact, any stressor that disrupts the quenching of
light energy may lead to bleaching (Wooldridge 2010; Baker and Cunning
2015). This is because the proximate cause of photo-oxidative stress is
excess excitation energy, but the upstream events that lead to this
situation may be diverse: indeed, elevated temperature may inhibit
Rubisco functioning (Jones et al. 1998) and the repair of the D1 protein
in photosystem II (Warner, Fitt, and Schmidt 1999), which reduces the
capacity of photochemical quenching and leads to an excess of light
energy. In this way, elevated temperature serves to reduce the threshold
above which light stresses the system (Hoegh-Guldberg 1999);
importantly, light is still the proximate stressor. Therefore, we
omitted temperature from the model to maintain a desired level of
simplicity, while still allowing photooxidative stress and bleaching to
be simulated with biological realism in response to light.

@@

Table 1. Model fluxes (mass-specific)

\begin{longtable}[c]{@{}llll@{}}

@@

\begin{minipage}[b]{0.10\columnwidth}\raggedright\strut
Symbol
\strut\end{minipage} &
\begin{minipage}[b]{0.48\columnwidth}\raggedright\strut
Description
\strut\end{minipage} &
\begin{minipage}[b]{0.25\columnwidth}\raggedright\strut
Units
\strut\end{minipage} &
\begin{minipage}[b]{0.10\columnwidth}\raggedright\strut
Eq. no.
\strut\end{minipage}\tabularnewline

@@

\begin{minipage}[t]{0.10\columnwidth}\raggedright\strut
\(j_X\)
\strut\end{minipage} &
\begin{minipage}[t]{0.48\columnwidth}\raggedright\strut
Prey uptake rate
\strut\end{minipage} &
\begin{minipage}[t]{0.25\columnwidth}\raggedright\strut
molX CmolH\textsuperscript{-1} d\textsuperscript{-1}
\strut\end{minipage} &
\begin{minipage}[t]{0.10\columnwidth}\raggedright\strut
3
\strut\end{minipage}\tabularnewline
\begin{minipage}[t]{0.10\columnwidth}\raggedright\strut
\(j_N\)
\strut\end{minipage} &
\begin{minipage}[t]{0.48\columnwidth}\raggedright\strut
Nitrogen uptake rate
\strut\end{minipage} &
\begin{minipage}[t]{0.25\columnwidth}\raggedright\strut
molN CmolH\textsuperscript{-1} d\textsuperscript{-1}
\strut\end{minipage} &
\begin{minipage}[t]{0.10\columnwidth}\raggedright\strut
4
\strut\end{minipage}\tabularnewline
\begin{minipage}[t]{0.10\columnwidth}\raggedright\strut
\(j_{HG}\)
\strut\end{minipage} &
\begin{minipage}[t]{0.48\columnwidth}\raggedright\strut
Host biomass formation rate
\strut\end{minipage} &
\begin{minipage}[t]{0.25\columnwidth}\raggedright\strut
CmolH CmolH\textsuperscript{-1} d\textsuperscript{-1}
\strut\end{minipage} &
\begin{minipage}[t]{0.10\columnwidth}\raggedright\strut
5
\strut\end{minipage}\tabularnewline
\begin{minipage}[t]{0.10\columnwidth}\raggedright\strut
\(r_{NH}\)
\strut\end{minipage} &
\begin{minipage}[t]{0.48\columnwidth}\raggedright\strut
Recycled nitrogen from host turnover
\strut\end{minipage} &
\begin{minipage}[t]{0.25\columnwidth}\raggedright\strut
molN CmolH\textsuperscript{-1} d\textsuperscript{-1}
\strut\end{minipage} &
\begin{minipage}[t]{0.10\columnwidth}\raggedright\strut
6
\strut\end{minipage}\tabularnewline
\begin{minipage}[t]{0.10\columnwidth}\raggedright\strut
\(\rho_N\)
\strut\end{minipage} &
\begin{minipage}[t]{0.48\columnwidth}\raggedright\strut
Nitrogen shared with the symbiont
\strut\end{minipage} &
\begin{minipage}[t]{0.25\columnwidth}\raggedright\strut
molN CmolH\textsuperscript{-1} d\textsuperscript{-1}
\strut\end{minipage} &
\begin{minipage}[t]{0.10\columnwidth}\raggedright\strut
7
\strut\end{minipage}\tabularnewline
\begin{minipage}[t]{0.10\columnwidth}\raggedright\strut
\(j_{eC}\)
\strut\end{minipage} &
\begin{minipage}[t]{0.48\columnwidth}\raggedright\strut
Excess carbon used to activate host CCMs
\strut\end{minipage} &
\begin{minipage}[t]{0.25\columnwidth}\raggedright\strut
molC CmolH\textsuperscript{-1} d\textsuperscript{-1}
\strut\end{minipage} &
\begin{minipage}[t]{0.10\columnwidth}\raggedright\strut
8
\strut\end{minipage}\tabularnewline
\begin{minipage}[t]{0.10\columnwidth}\raggedright\strut
\(j_{CO_2}\)
\strut\end{minipage} &
\begin{minipage}[t]{0.48\columnwidth}\raggedright\strut
CO\textsubscript{2} input to photosynthesis
\strut\end{minipage} &
\begin{minipage}[t]{0.25\columnwidth}\raggedright\strut
molCO\textsubscript{2} CmolH\textsuperscript{-1} d\textsuperscript{-1}
\strut\end{minipage} &
\begin{minipage}[t]{0.10\columnwidth}\raggedright\strut
9
\strut\end{minipage}\tabularnewline
\begin{minipage}[t]{0.10\columnwidth}\raggedright\strut
\(j_L\)
\strut\end{minipage} &
\begin{minipage}[t]{0.48\columnwidth}\raggedright\strut
Light absorption rate
\strut\end{minipage} &
\begin{minipage}[t]{0.25\columnwidth}\raggedright\strut
molph CmolS\textsuperscript{-1} d\textsuperscript{-1}
\strut\end{minipage} &
\begin{minipage}[t]{0.10\columnwidth}\raggedright\strut
11
\strut\end{minipage}\tabularnewline
\begin{minipage}[t]{0.10\columnwidth}\raggedright\strut
\(r_{CH}\)
\strut\end{minipage} &
\begin{minipage}[t]{0.48\columnwidth}\raggedright\strut
Recycled CO\textsubscript{2} from host turnover
\strut\end{minipage} &
\begin{minipage}[t]{0.25\columnwidth}\raggedright\strut
molCO\textsubscript{2} CmolH\textsuperscript{-1} d\textsuperscript{-1}
\strut\end{minipage} &
\begin{minipage}[t]{0.10\columnwidth}\raggedright\strut
12
\strut\end{minipage}\tabularnewline
\begin{minipage}[t]{0.10\columnwidth}\raggedright\strut
\(r_{CS}\)
\strut\end{minipage} &
\begin{minipage}[t]{0.48\columnwidth}\raggedright\strut
Recycled CO\textsubscript{2} from symbiont turnover
\strut\end{minipage} &
\begin{minipage}[t]{0.25\columnwidth}\raggedright\strut
molCO\textsubscript{2} CmolS\textsuperscript{-1} d\textsuperscript{-1}
\strut\end{minipage} &

>>>>

\usepackage[T1]{fontenc}
\usepackage{lmodern}
\usepackage{amssymb,amsmath}
\usepackage{ifxetex,ifluatex}
\usepackage{fixltx2e} % provides \textsubscript
% use upquote if available, for straight quotes in verbatim environments
\IfFileExists{upquote.sty}{\usepackage{upquote}}{}
\ifnum 0\ifxetex 1\fi\ifluatex 1\fi=0 % if pdftex
  \usepackage[utf8]{inputenc}
\else % if luatex or xelatex
  \usepackage{fontspec}
  \ifxetex
    \usepackage{xltxtra,xunicode}
  \fi
  \defaultfontfeatures{Mapping=tex-text,Scale=MatchLowercase}
  \newcommand{\euro}{€}
\fi
% use microtype if available
\IfFileExists{microtype.sty}{\usepackage{microtype}}{}
\usepackage{longtable}
\usepackage{graphicx}
% We will generate all images so they have a width \maxwidth. This means
% that they will get their normal width if they fit onto the page, but
% are scaled down if they would overflow the margins.
\makeatletter
\def\maxwidth{\ifdim\Gin@nat@width>\linewidth\linewidth
\else\Gin@nat@width\fi}
\makeatother
\let\Oldincludegraphics\includegraphics
\renewcommand{\includegraphics}[1]{\Oldincludegraphics[width=\maxwidth]{#1}}
\ifxetex
  \usepackage[setpagesize=false, % page size defined by xetex
              unicode=false, % unicode breaks when used with xetex
              xetex]{hyperref}
\else
  \usepackage[unicode=true]{hyperref}
\fi
\hypersetup{breaklinks=true,
            bookmarks=true,
            pdfauthor={},
            pdftitle={A dynamic bioenergetic model for coral-Symbiodinium symbioses and coral bleaching as an alternate stable state},
            colorlinks=true,
            urlcolor=blue,
            linkcolor=magenta,
            pdfborder={0 0 0}}
\urlstyle{same}  % don't use monospace font for urls
\setlength{\parindent}{0pt}
\setlength{\parskip}{6pt plus 2pt minus 1pt}
\setlength{\emergencystretch}{3em}  % prevent overfull lines
\setcounter{secnumdepth}{0}
% Pandoc toggle for numbering sections (defaults to be off)
\setcounter{secnumdepth}{0}
% Pandoc header

@@

  \title{A dynamic bioenergetic model for coral-\emph{Symbiodinium} symbioses and
coral bleaching as an alternate stable state}
    \author[University of Hawaii]{Ross Cunning\corref{c1}}
   \ead{ross.cunning@gmail.com} 
   \cortext[c1]{Corresponding Author}
    \author[UCSB, MSI]{Erik B. Muller}

    \author[HIMB]{Ruth D. Gates}

    \author[UCSB]{Roger M. Nisbet}

      \address[HIMB]{Hawaii Institute of Marine Biology, School of Ocean and Earth Science
and Technology, University of Hawaii at Manoa, USA}
    \address[UCSB]{Department of Ecology, Evolution, and Marine Biology, University of
California, Santa Barbara, USA}
    \address[MSI]{Marine Science Institute, University of California, Santa Barbara, USA}

  \begin{abstract}
  Coral reef ecosystems owe their ecological success--and vulnerability to
  climate change--to the symbiotic metabolism of corals and
  \emph{Symbiodinium} spp. The urgency to understand and predict the
  stability and breakdown of these symbioses (i.e., coral `bleaching')
  demands the development and application of theoretical tools. Here, we
  develop a dynamic bioenergetic model of coral-\emph{Symbiodinium}
  symbioses that demonstrates realistic steady-state patterns in coral
  growth and symbiont abundance across gradients of light, nutrients, and
  feeding. Furthermore, by including a mechanistic treatment of
  photo-oxidative stress, the model displays dynamics of bleaching and
  recovery that can be explained as transitions between alternate stable
  states. These dynamics reveal that `healthy' and `bleached' states
  correspond broadly to nitrogen- and carbon-limitation of the system,
  with transitions between them occurring as integrated responses to
  multiple environmental factors. Indeed, a suite of complex emergent
  behaviors reproduced by the model (e.g., bleaching is exacerbated by
  nutrients and attenuated by feeding) suggests it unifyingly captures
  many important attributes of the system; meanwhile, its modular
  framework and open source R code are designed to facilitate further
  problem-specific development. We see significant potential for this
  modeling framework to generate testable hypotheses and predict
  integrated, mechanistic responses of corals to environmental change,
  with important implications for understanding the performance and
  maintenance of symbiotic systems.
  \end{abstract}

@@

In reef coral symbioses, intracellular \emph{Symbiodinium} translocate
photosynthetically fixed carbon to support coral metabolism, while the
animal host provides access to inorganic nutrients and carbon dioxide
(Muscatine and Porter 1977). Previous application of DEB theory to this
syntrophic system (Muller et al. 2009) demonstrated a stable symbiotic
relationship and qualitatively realistic growth and biomass ratios
across gradients of ambient irradiance, nutrients, and food. This model
(as well as the present work) assumes that 1) \emph{Symbiodinium} has
priority access to fixed carbon through photosynthesis, 2) the coral
animal has priority access to nitrogen through contact with seawater and
feeding on prey, and 3) each partner shares with the other only what it
cannot use for its own growth. In its simplest form, this principle of
sharing the surplus is sufficient to describe the dynamics of diverse
syntrophic organs and organisms (e.g., trees, duckweeds, corals),
suggesting the mechanism is mathematically and evolutionarily robust
(Nisbet et al., in prep.).

@@

\begin{enumerate}
\def\labelenumi{\arabic{enumi}.}
\item
  \emph{Develop a new module of photo-oxidative stress.} Of primary
  interest to coral biologists and ecologists is symbiosis dysfunction
  under environmental stress, resulting in coral ``bleaching''--the loss
  of algal symbionts from the association (Jokiel and Coles 1977).
  Bleaching is thought to be triggered by photo-oxidative stress in
  \emph{Symbiodinium} (Weis 2008), which has been modeled previously as
  a response to absolute external irradiance (Eynaud, Nisbet, and Muller
  2011). However, this response may also depend on self-shading by other
  symbionts, CO\textsubscript{2} availability at the site of
  photosynthesis, and other non-photochemical quenching mechanisms. We
  incorporate these features into a new photo-oxidative stress module
  linking overreduction of the photosynthetic light reactions to
  downstream impacts of photoinhibition and photodamage. Importantly,
  this formulation introduces a link between
  CO\textsubscript{2}-limitation of photosynthesis and bleaching, and
  potential synergistic roles for heterotrophy and nutrient availability
  in influencing bleaching responses (Wooldridge 2014b)
\item
  \emph{Reduce theoretical and mathematical complexity.} Following the
  logic of Jager, Martin, and Zimmer (2013), we exclude certain features
  of formal DEB theory in order to capture behaviors of interest with
  the simplest possible formulation. Here, we present a model without
  reserves, maturity, or reproduction (see Kooijman 2010). This
  formulation precludes modeling the full life cycle of corals as
  reproduction, larval stages, and metamorphosis are not considered, but
  greatly reduces theoretical complexity and parameter numbers, which is
  advantageous given the relative paucity of data for corals. Moreover,
  dynamics of the symbiosis (i.e., changes in symbiont to host biomass,
  including bleaching and recovery) and coral biomass growth are
  efficiently captured with this simpler formulation, which also
  increases accessibility for biologists and ecologists without
  requiring significant expertise in DEB theory.
\item
  \emph{Provide well-documented, open-access code.} In order to
  facilitate the continued development and application of theoretical
  modeling tools for coral symbioses, we provide open access to the
  model in the form of detailed, commented code written in the R
  language (R Core Team 2014). With an accessible and modular framework,
  we envision this model as a resource for further development by the
  scientific community to include additional complexity and
  problem-specific components. We chose the R language, which is freely
  available and in common use by biologists and ecologists, to reach the
  widest possible audience with this work.
\end{enumerate}

@@

In this dynamical system, both partners acquire and use carbon and
nitrogen to construct biomass: the symbiont fixes carbon through
photosynthesis, and receives nitrogen shared by the host, while the host
acquires nitrogen from the environment and receives carbon shared by the
symbiont. A graphical representation of the model is presented in Fig.
1, and each model flux and parameter is defined in Tables 1 and 2,
respectively. We use C-moles (C-mol) as the unit of biomass for
consistency with the rigorous mass balance of DEB theory: 1 C-mol is
equivalent to the amount of biomass containing 1 mole of carbon atoms.
Host biomass (expressed as C-mol H), symbiont biomass (expressed as
C-mol S), and prey biomass (expressed as C-mol X) have fixed, but
different, molar N:C ratios (Table 2). Biomass is produced from carbon
and nitrogen by synthesizing units (SU), which are mathematical
specifications of the formation of a product from two substrates; we use
the ``parallel complementary'' formulation of Kooijman (2010) (page 105,
Fig. 3.7) to specify these fluxes. The two state variables of this
system are symbiont biomass and coral biomass; because resources are
acquired proportionally to surface area, and surface area is
proportional to volume, biomass increases exponentially during growth
(indeed, corals grow exponentially (Bak 1976)).

Environmental stress is implemented in the form of photooxidative
stress, which is thought to be a primary trigger of coral bleaching
(Lesser 1997; Weis 2008; Wooldridge 2009). To simulate bleaching, we
model the absorption and quenching of light energy by photochemistry and
non-photochemical quenching, and the responses that occur (i.e.,
photoinhibition, photodamage, and symbiont loss) when these quenching
capacities are overwhelmed. While bleaching in response to high light
alone has been observed experimentally (Schutter et al. 2011; Downs et
al. 2013), mass coral bleaching events occur concurrently with high
temperature (Hoegh-Guldberg 1999); thus, it is important to justify our
consideration of light as the primary stressor. In reality, light and
temperature interact synergistically (Coles and Jokiel 1978; Jones et
al. 1998), and in fact, any stressor that disrupts the quenching of
light energy may lead to bleaching (Wooldridge 2010; Baker and Cunning
2015). This is because the proximate cause of photo-oxidative stress is
excess excitation energy, but the upstream events that lead to this
situation may be diverse: indeed, elevated temperature may inhibit
Rubisco functioning (Jones et al. 1998) and the repair of the D1 protein
in photosystem II (Warner, Fitt, and Schmidt 1999), which reduces the
capacity of photochemical quenching and leads to an excess of light
energy. In this way, elevated temperature serves to reduce the threshold
above which light stresses the system (Hoegh-Guldberg 1999);
importantly, light is still the proximate stressor. Therefore, we
omitted temperature from the model to maintain a desired level of
simplicity, while still allowing photooxidative stress and bleaching to
be simulated with biological realism in response to light.

@@

Table 1. Model fluxes (mass-specific). Units are explained in the text.

\begin{longtable}[]{@{}llll@{}}
\toprule
Symbol & Description & Units & Eq. no.\tabularnewline
\midrule
\endhead
\(j_X\) & Prey uptake rate & mol X C-mol H\textsuperscript{-1}
d\textsuperscript{-1} & 3\tabularnewline
\(j_N\) & Nitrogen uptake rate & mol N Cmol H\textsuperscript{-1}
d\textsuperscript{-1} & 4\tabularnewline
\(j_{HG}\) & Host biomass formation rate & C-mol H C-mol
H\textsuperscript{-1} d\textsuperscript{-1} & 5\tabularnewline
\(j_{HT}\) & Host biomass turnover rate & C-mol H C-mol
H\textsuperscript{-1} d\textsuperscript{-1} & 21\tabularnewline
\(r_{NH}\) & Recycled nitrogen from host turnover & mol N C-mol
H\textsuperscript{-1} d\textsuperscript{-1} & 6\tabularnewline
\(\rho_N\) & Nitrogen shared with the symbiont & mol N C-mol
H\textsuperscript{-1} d\textsuperscript{-1} & 7\tabularnewline
\(j_{eC}\) & Excess carbon used to activate host CCMs & mol C C-mol
H\textsuperscript{-1} d\textsuperscript{-1} & 8\tabularnewline
\(j_{CO_2}\) & CO\textsubscript{2} input to photosynthesis & mol
CO\textsubscript{2} C-mol H\textsuperscript{-1} d\textsuperscript{-1} &
9\tabularnewline
\(j_L\) & Light absorption rate & mol q C-mol S\textsuperscript{-1}
d\textsuperscript{-1} & 11\tabularnewline
\(r_{CH}\) & Recycled CO\textsubscript{2} from host turnover & mol
CO\textsubscript{2} C-mol H\textsuperscript{-1} d\textsuperscript{-1} &
12\tabularnewline
\(r_{CS}\) & Recycled CO\textsubscript{2} from symbiont turnover & mol
CO\textsubscript{2} C-mol S\textsuperscript{-1} d\textsuperscript{-1} &
13\tabularnewline
\(j_{CP}\) & Photosynthesis rate & mol C C-mol S\textsuperscript{-1}
d\textsuperscript{-1} & 14\tabularnewline
\(j_{eL}\) & Light energy in excess of photochemistry & mol q C-mol
S\textsuperscript{-1} d\textsuperscript{-1} & 15\tabularnewline
\(j_{NPQ}\) & Total capacity of NPQ & mol q C-mol S\textsuperscript{-1}
d\textsuperscript{-1} & 16\tabularnewline
\(c_{ROS}\) & ROS production proportional to baseline & -- &
17\tabularnewline
\(r_{NS}\) & Recycled nitrogen from symbiont turnover & mol N C-mol
S\textsuperscript{-1} d\textsuperscript{-1} & 18\tabularnewline
\(j_{SG}\) & Symbiont biomass formation rate & C-mol S C-mol
S\textsuperscript{-1} d\textsuperscript{-1} & 19\tabularnewline
\(\rho_C\) & Fixed carbon shared with host & mol C C-mol
S\textsuperscript{-1} d\textsuperscript{-1} & 20\tabularnewline
\(j_{ST}\) & Symbiont biomass turnover rate & C-mol S C-mol
S\textsuperscript{-1} d\textsuperscript{-1} & 21\tabularnewline
\bottomrule
\end{longtable}

Table 2. Model parameters (units explained in the text). Derivations of
each parameter value along with supporting references are provided in
the Supplementary Information.

\begin{longtable}[]{@{}llll@{}}

@@

\begin{minipage}[b]{0.09\columnwidth}\raggedright\strut
Symbol\strut
\end{minipage} & \begin{minipage}[b]{0.46\columnwidth}\raggedright\strut
Description\strut
\end{minipage} & \begin{minipage}[b]{0.24\columnwidth}\raggedright\strut
Units\strut
\end{minipage} & \begin{minipage}[b]{0.09\columnwidth}\raggedright\strut
Value\strut
\end{minipage}\tabularnewline

@@

\begin{minipage}[t]{0.09\columnwidth}\raggedright\strut
\(n_{NH}\)\strut
\end{minipage} & \begin{minipage}[t]{0.46\columnwidth}\raggedright\strut
N:C molar ratio in host biomass\strut
\end{minipage} & \begin{minipage}[t]{0.24\columnwidth}\raggedright\strut
--\strut
\end{minipage} & \begin{minipage}[t]{0.09\columnwidth}\raggedright\strut
\(0.18\)\strut
\end{minipage}\tabularnewline
\begin{minipage}[t]{0.09\columnwidth}\raggedright\strut
\(n_{NS}\)\strut
\end{minipage} & \begin{minipage}[t]{0.46\columnwidth}\raggedright\strut
N:C molar ratio in symbiont biomass\strut
\end{minipage} & \begin{minipage}[t]{0.24\columnwidth}\raggedright\strut
--\strut
\end{minipage} & \begin{minipage}[t]{0.09\columnwidth}\raggedright\strut
\(0.13\)\strut
\end{minipage}\tabularnewline
\begin{minipage}[t]{0.09\columnwidth}\raggedright\strut
\(n_{NX}\)\strut
\end{minipage} & \begin{minipage}[t]{0.46\columnwidth}\raggedright\strut
N:C molar ratio in prey biomass\strut
\end{minipage} & \begin{minipage}[t]{0.24\columnwidth}\raggedright\strut
--\strut
\end{minipage} & \begin{minipage}[t]{0.09\columnwidth}\raggedright\strut
\(0.2\)\strut
\end{minipage}\tabularnewline
\begin{minipage}[t]{0.09\columnwidth}\raggedright\strut
\(j_{HT}^0\)\strut
\end{minipage} & \begin{minipage}[t]{0.46\columnwidth}\raggedright\strut
Specific maintenance rate of host biomass\strut
\end{minipage} & \begin{minipage}[t]{0.24\columnwidth}\raggedright\strut
C-mol H C-mol H\textsuperscript{-1} d\textsuperscript{-1}\strut
\end{minipage} & \begin{minipage}[t]{0.09\columnwidth}\raggedright\strut
\(0.03\)\strut
\end{minipage}\tabularnewline
\begin{minipage}[t]{0.09\columnwidth}\raggedright\strut
\(j_{ST}^0\)\strut
\end{minipage} & \begin{minipage}[t]{0.46\columnwidth}\raggedright\strut
Specific maintenance rate of symbiont biomass\strut
\end{minipage} & \begin{minipage}[t]{0.24\columnwidth}\raggedright\strut
C-mol S C-mol S\textsuperscript{-1} d\textsuperscript{-1}\strut
\end{minipage} & \begin{minipage}[t]{0.09\columnwidth}\raggedright\strut
\(0.03\)\strut
\end{minipage}\tabularnewline
\begin{minipage}[t]{0.09\columnwidth}\raggedright\strut
\(\sigma_{NH}\)\strut
\end{minipage} & \begin{minipage}[t]{0.46\columnwidth}\raggedright\strut
Proportion host N-turnover recycled to host\strut
\end{minipage} & \begin{minipage}[t]{0.24\columnwidth}\raggedright\strut
--\strut
\end{minipage} & \begin{minipage}[t]{0.09\columnwidth}\raggedright\strut
\(0.9\)\strut
\end{minipage}\tabularnewline
\begin{minipage}[t]{0.09\columnwidth}\raggedright\strut
\(\sigma_{CH}\)\strut
\end{minipage} & \begin{minipage}[t]{0.46\columnwidth}\raggedright\strut
Proportion host C-turnover recycled to photosynthesis\strut
\end{minipage} & \begin{minipage}[t]{0.24\columnwidth}\raggedright\strut
--\strut
\end{minipage} & \begin{minipage}[t]{0.09\columnwidth}\raggedright\strut
\(0.1\)\strut
\end{minipage}\tabularnewline
\begin{minipage}[t]{0.09\columnwidth}\raggedright\strut
\(\sigma_{NS}\)\strut
\end{minipage} & \begin{minipage}[t]{0.46\columnwidth}\raggedright\strut
Proportion symbiont N-turnover recycled to symbiont\strut
\end{minipage} & \begin{minipage}[t]{0.24\columnwidth}\raggedright\strut
--\strut
\end{minipage} & \begin{minipage}[t]{0.09\columnwidth}\raggedright\strut
\(0.9\)\strut
\end{minipage}\tabularnewline
\begin{minipage}[t]{0.09\columnwidth}\raggedright\strut
\(\sigma_{CS}\)\strut
\end{minipage} & \begin{minipage}[t]{0.46\columnwidth}\raggedright\strut
Proportion symbiont C-turnover recycled to photosynthesis\strut
\end{minipage} & \begin{minipage}[t]{0.24\columnwidth}\raggedright\strut
--\strut
\end{minipage} & \begin{minipage}[t]{0.09\columnwidth}\raggedright\strut
\(0.9\)\strut
\end{minipage}\tabularnewline
\begin{minipage}[t]{0.09\columnwidth}\raggedright\strut
\(j_{Xm}\)\strut
\end{minipage} & \begin{minipage}[t]{0.46\columnwidth}\raggedright\strut
Maximum specific feeding rate of host\strut
\end{minipage} & \begin{minipage}[t]{0.24\columnwidth}\raggedright\strut
molX C-mol H\textsuperscript{-1} d\textsuperscript{-1}\strut
\end{minipage} & \begin{minipage}[t]{0.09\columnwidth}\raggedright\strut
\(0.13\)\strut
\end{minipage}\tabularnewline
\begin{minipage}[t]{0.09\columnwidth}\raggedright\strut
\(K_X\)\strut
\end{minipage} & \begin{minipage}[t]{0.46\columnwidth}\raggedright\strut
Half-saturation constant for prey uptake by host\strut
